
\documentclass[10pt, oneside]{article}
\usepackage[letterpaper, scale=0.9, centering]{geometry}
\usepackage{graphicx}			
\usepackage{url,hyperref}
\usepackage{amssymb,amsmath,mathpartir}
\usepackage{currfile,xstring,multicol,array}
\usepackage[dvipsnames]{xcolor}
\usepackage[labelformat=empty]{caption}

\title{UCSD CSE131 F19 -- Diamondback}

\usepackage{listings}

\lstset{
  basicstyle=\ttfamily,
  mathescape,
  columns=fullflexible
}

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}
\begin{document}
\maketitle 

{\bf Due Date:} 11pm Wednesday, November 6 \hspace{2em} {\bf Closed to Collaboration}

You will implement Diamondback, a language with functions and a static type
system.

Classroom: \url{https://classroom.github.com/a/TyAjQxca} \hspace{1em} Github: \url{https://github.com/ucsd-cse131-f19/pa4-student}


\section*{Syntax}

The concrete syntax and type language for Diamondback is below. We use
$\cdots$ to indicate \textit{zero or more} of the previous
element\footnote{This is different than the last assignment, but important
for avoiding some notational clutter.}. So a program $p$ is a sequence of zero
or more definitions $d$ followed by an expression $e$, and a definition has
zero or more arguments $x \texttt{:}\tau$ and one or more expressions in its
body $e \cdots$. There are \fbox{boxes} around the new pieces of concrete
syntax.

\[
\begin{array}{ll}
\begin{array}{lrl}
e & := & n \mid \texttt{true} \mid \texttt{false} \mid x \\
  & \mid  & \texttt{(let (($x\ e$) ($x\ e$) $\cdots$) $e\ e \cdots$)} \\
  & \mid  & \texttt{(if $e\ e \ e$)} \\
  & \mid  & \texttt{($op_2$ $e\ e$)} \mid \texttt{($op_1$ $e$)} \\
  & \mid  & \texttt{(while $e\ e\ e \cdots$)} \mid \texttt{(set $x\ e$)} \\
  & \mid  & \fbox{\texttt{($f\ e\cdots$)}} \\
d & := & \fbox{\texttt{(def $f$ ($x\ $:$\ \tau \cdots$) :$\ \tau$ $e\ e \cdots$)}} \\
p & := & \fbox{$d \cdots e$} \\
op_1 & := & \texttt{add1} \mid \texttt{sub1} \mid \texttt{isNum} \mid \texttt{isBool} \mid \fbox{\texttt{print}} \\
op_2 & := & \texttt{+} \mid \texttt{-} \mid \texttt{*} \mid \texttt{<} \mid \texttt{>} \mid \texttt{==} \\
n & := & \textrm{63-bit signed number literals} \\
x,f & := & \textrm{variable and function names} \\
\end{array}
&
\begin{array}{lrl}
\tau & := & \texttt{Num} \mid \texttt{Bool} \\
\Delta & := & \{f: \tau \cdots \rightarrow \tau, \cdots\} \\
\Delta[f] & \textit{means} & \textit{look up the type of } f \textit{ in } \Delta \\
\Gamma & := & \{x:\tau,\cdots\} \\
\Gamma[x] & \textit{means} & \textit{look up the type of } x \textit{ in } \Gamma \\
(x,\tau)::\Gamma & \textit{means} & \textit{add } x \textit{ to } \Gamma \textit{ with type } \tau \\
\Delta;\Gamma \vdash e : \tau & \textit{means} & \textit{with definitions } \Delta \textit{ and env } \Gamma, e \textit{ has type } \tau \\
\Delta \vdash_d d : \checkmark & \textit{means} & \textit{with definitions } \Delta \textit{ the definition } d \textit{ type-checks} \\
\vdash_p p : \checkmark & \textit{means} & \textit{the program } p \textit{ type checks } \\
\end{array}
\end{array}
\]

An example program that computes whether a number is even or odd extremely
inefficiently and prints several examples is:

\begin{tabular}{lr}
\begin{lstlisting}
(def even (n : Num) : Bool
  (if (== n 0) true (odd (- n 1))))
(def odd (n : Num) : Bool
  (if (== n 0) false (even (- n 1))))
(def test() : Bool
  (print (even 30))
  (print (odd 30))
  (print (even 57))
  (print (odd 57)))

(test)
\end{lstlisting}
& \hspace{5em}
\begin{lstlisting}
Expected output:
true
false
false
true
true
\end{lstlisting}
\end{tabular}

Another example program that defines a few small functions and uses them is

\begin{tabular}{lr}
\begin{lstlisting}
(def manhattan (x1 : Num y1 : Num x2 : Num y2 : Num) : Num
  (+ (abs (- x1 x2)) (abs (- y1 y2))))
(def abs (a : Num) : Num
  (if (< a 0) (* a -1) a))
(def printmultiples (a : Num b : Num) : Bool
  (let ((orig (+ a 1)))
    (while (> a 0)
      (print (* (- orig a) b))
      (set a (- a 1)))))

(let ((ignore 0))
  (print (abs -3))
  (print (manhattan 0 0 3 4))
  (print (manhattan 3 4 0 0))
  (printmultiples 3 2))
\end{lstlisting}
& \hspace{5em}
\begin{lstlisting}
Expected output:
3
7
7
2
4
6
false
\end{lstlisting}
\end{tabular}

\section*{Semantics}

\subsection*{Function Definitions and Applications}

The main new feature in Diamondback is function definitions $d$ and function
applications $\texttt{($f\ e\cdots$)}$. A function application $\texttt{($f\
e_a\cdots$)}$ uses the function definition with the matching name
$\texttt{(def $f$ ($x$ :$\tau \cdots$) :$\ \tau\ e \cdots$)}$, and should evaluate to
the same result as $\texttt{(let (($x\ e_b$) $\cdots$) $e \cdots$)}$, where
the argument expressions $e_a\cdots$ come from the application, the names
$x\cdots$ come from the definition, and the body expressions $e_b\cdots$ come
from the definition.\footnote{Note that replacing application expressions
with let expressions is \emph{not} a strategy that works in general in the
compiler, because the body expressions $e_b\cdots$ could contain other
function applications. This description is, however, a useful way to describe
the behavior of a function application succinctly and is a perfectly valid
way to evaluate functions ``by hand'' when we can write out all the
intermediate steps with concrete values. See the reading for more detail:
\url{https://ucsd-cse131-f19.github.io/lectures/10-22-lec8/notes.pdf}}\footnote{We
actually need to be a little bit careful here. While this rule worked fine
for the single-argument functions in the reading, here we'd really want a
version of let that doesn't include earlier bindings in later ones to avoid
clashes between names in scope and names in the function's argument list.
This is really only relevant if you're writing things out on paper, and
doesn't affect the design of a calling convention at all.}

\subsection*{The Main Expression}

Diamondback programs are expected to have a single expression after the
definition list that is the main entry point for the program. This expression
has {\tt input} bound to the user's input by default.

As an example, consider this program:

\begin{lstlisting}
(def (abs x : Num) : Num
  (if (< x 0) (* -1 x) x))
(* (abs input) 2)
\end{lstlisting}

We could think of its evaluation taking these steps (with an input of 5):

\begin{lstlisting}
$\rightarrow$ (* (abs input) 2)
$\rightarrow$ (* (abs 5) 2)
$\rightarrow$ (* (let ((x 5)) (if (< x 0) (* -1 x) x)) 2)
$\rightarrow$ (* (if (< 5 0) (* -1 5) 5) 2)
$\rightarrow$ (* (if false (* -1 5) 5) 2)
$\rightarrow$ (* 5 2)
$\rightarrow$ 10
\end{lstlisting}

\subsection*{Printing}

Diamondback also adds a new primitive, {\tt print}, that prints a value to
the console followed by a newline. Numbers should print as their
user-interpreted value, so {\tt (print 22)} should print {\tt 22}, and {\tt
(print true)} should print {\tt true}.

The entire {\tt print} expression evaluates to the same value as its argument
(which is the value that gets printed), so {\tt (print (+ 1 7))} evaluates to
{\tt 8} in addition to printing it.

\section*{Type Checking}

Diamondback has essentially the same type rules as Copperhead for
expressions, with two changes. First, all of the rules contain a definitions
environment $\Delta$ in addition to the type environment $\Gamma$. Second,
there are two new rules:

\begin{mathpar}
\inferrule*[Left=TR-Print]
{\Gamma \vdash e : \tau}
{\Delta;\Gamma \vdash \texttt{(print $e$)} : \tau}
\and
\inferrule*[Left=TR-App]
{\Delta[f] = \tau_1 \cdots \tau_n \rightarrow \tau_r \and \left(\Delta;\Gamma \vdash e_1 : \tau_1 \right) \cdots \left(\Delta;\Gamma \vdash e_n : \tau_n \right)}
{\Delta;\Gamma \vdash \texttt{($f$ $e_1 \cdots e_n$)} : \tau_r}
\end{mathpar}

TR-Print says that {\tt print} expressions' result is the same type $\tau$ as
the argument, which matches the semantics.

In English, TR-App rule says

\begin{quote}
If the function definition $f$ has argument types $\tau_1 \cdots$ and return type $\tau_r$, and the
arguments $e_1\cdots$ of an application of $f$ have matching types, then the application has type $\tau_r$.
\end{quote}

It's common to write the type of definitions as $\tau_1 \cdots \rightarrow
\tau_r$, also called an ``arrow type'', to be evocative of taking a number of
argument types and producing a result type.

The existing rules are unchanged aside from tracking $\Delta$ (which only
TR-App uses).

There are also two new rules, one for type-checking definitions, and one for
type-checking programs. They use slightly different $\vdash$ notation with
subscripts $_d$ and $_p$ to indicate that they have meaning for pieces of
syntax other than expressions. Since, unlike expressions, we don't calculate
their overall type, we simply use $\checkmark$ in the rule to indicate that
they pass all checks.

\begin{mathpar}
\inferrule*[Left=TR-Def]
{\Delta; \{x_1:\tau_1, \cdots x_n:\tau_n\} \vdash e \cdots : \tau_r}
{\Delta \vdash_d \texttt{(def $f$ ($x_1$:$\tau_1 \cdots x_n$:$\tau_n$):$\tau_r$ $e \cdots$)} : \checkmark}
\\
\inferrule*[Left=TR-Prog]
{(\Delta; \vdash_d d_1 : \checkmark)\cdots(\Delta; \vdash_d d_n : \checkmark) \and \Delta; \{{\tt input}:{\tt Num}\} \vdash e : \tau}
{\vdash_p d_1 \cdots d_n e : \checkmark}
\end{mathpar}

Where in $\vdash_p$, $\Delta$ is constructed by mapping each definition name
to an arrow type made of its argument types and return type, so 
\texttt{(def $f$ ($x$:$\tau_1 \cdots x_n$:$\tau_n$):$\tau_r$ $e \cdots$)} would 
appear in $\Delta$ as $\{f: \tau_1\cdots\tau_n \rightarrow \tau_r\}$. In English,
TR-Def says that a definition type-checks if its body has the expected return
type $\tau_r$ when type-checked in an environment with just the arguments of
the definition mapped to their declared types. A program $p$ type-checks if
all of its definitions type-check and its main expression has some type in
the environment that assumes {\tt input} has type {\tt Num} (along with also
assuming the declared definitions).

\section*{New Errors \& Miscellaneous}

Type errors should be reported with {\tt "Type mismatch"} as usual, including
type errors resulting from the new rules, including passing the wrong number
of arguments to a function. If a function application uses a function name
that isn't defined, the compiler should report an error containing {\tt
"Unbound"}.

It's allowed for variables and functions to use the same name, so there could
be a top-level definition named {\tt f} and a variable in an argument or let
named {\tt f}.

It's a well-formedness error for multiple functions to have the same name, or
for multiple arguments within the same function to have the same name. Report
these cases with an error that contains the string {\tt "Multiple functions"}
and {\tt "Multiple bindings"} respectively.

An empty function body should be reported with {\tt "Invalid"} as with other
syntax errors.

\section*{Implementation Recommendations and Details}

\subsection*{Registers Used by {\tt main}}

You may have reasons to want to use registers like {\tt rbx, rbp, rdi} or
others. The assembly generated by gcc and clang for {\tt main} may use these
registers as well, so they should be saved at the beginning of {\tt
our\_code\_starts\_here} and restored before the final {\tt ret}. You can use
{\tt push rbx} and {\tt pop rbx} to accomplish this.

\subsection*{Stack Alignment}

Some systems require that the stack pointer {\tt rsp} be aligned at a 16-byte
(2 word) boundary before making calls into library functions that use system
calls, like {\tt printf}. If you get stack alignment segmentation
faults\footnote{We saw this in class
\url{https://github.com/ucsd-cse131-f19/ucsd-cse131-f19.github.io/blob/master/lectures/10-10-lec5/compile.ml\#L106}}
you may want to make sure your calling convention always moves {\tt rsp} by
multiples of 16, which could mean leaving an extra word of space on some
calls. For example, if your calling convention uses 2 words for the old value
of {\tt rsp} and the return address, then a function call with an {\it odd}
number of arguments could end up on an 8-byte boundary. You can test this by
using {\tt print} in functions with varying numbers of arguments.

\subsection*{Moving Labels into Memory}

In class, we used code like {\tt mov [rsp-16], after\_call} to move a label
into memory. This actually requires two instructions on some platforms. If
you see an error like {\tt "format does not support 32-bit absolute
addresses."} you may be running into this. The solution is simple, just save
the label into a register first:

\begin{lstlisting}
mov rax, after_call
mov [rsp-16], rax
\end{lstlisting}

\subsection*{Print}

While you're free to implement {\tt print} in any way you prefer that works,
one that we found expedient is to call a function defined in {\tt main.c}.
This requires using C's calling convention. On x86-64, this means moving the
argument into register {\tt rdi}, moving {\tt rsp} to free space at the top
of the stack, and then using the {\tt call} instruction to push the current
code address to the stack and jump to the function you wrote in {\tt main}.
On return, {\tt rsp} should be moved back to its original location, and your
generated code should ensure that the printed value ends up in {\tt rax}.

Keep in mind that if you use registers other than {\tt rax}, they may be
overwritten during the use of print. Wikipedia has a reasonable, brief,
accurate summary of the callee-save vs. caller-save behavior at
\url{https://en.wikipedia.org/wiki/X86_calling_conventions\#System_V_AMD64_ABI}.
We assuming that users of Diamondback will use {\tt gcc} and {\tt clang} on
systems that use this convention, not the Microsoft convention.

\section*{Required Tests}

You {\bf must} write the following programs to test your compiler. You can
pick any little details you want like function names and base case behavior,
but they must be in files with these names:

\begin{itemize}
\item {\tt input/fibonacci.boa} -- A program that defines a fibonacci
function and calls it on {\tt input}.
\item {\tt input/remainder.boa} -- A program that defines a function that
takes two arguments and uses repeated subtraction to get the remainder of the
first divided by the second. Test it on several pairs of inputs and include
{\tt print} expressions demonstrating it. (Yes, we are asking you to write
tests for the function you're writing as a test.)
\item {\tt input/isprime.boa} -- A program that uses the definition from
{\tt remainder.boa} (you can copy/paste it) along with a loop to test if {\tt
input} is prime, and print {\tt true} if it is and {\tt false} if not. Your
algorithm does not have to be sophisticated; it's fine to try the
numbers from 2 to {\tt input} and check the remainder.
\item {\tt input/deepstack.boa} -- A program of your choice that has at least
3 different functions, each with a different number of arguments. When run,
it should require a runtime stack of at least depth 10, where each function
has at least one frame represented on the stack.
\end{itemize}

These tests together will be worth 20\% of your grade (you still get credit
even if your compiler can't fully run them as long as they are written).

You should start with smaller tests and write many more tests than these. We
provide these because they usefully exercise your compiler and demonstrate
that you can write some meaningful computations now.

In the assignment {\tt pa4-written} on Gradescope, include these 4 programs
in your submitted PDF.


\section*{Describing Your Calling Convention}

As you implement Diamondback, you will need to make a number of decisions,
not least of which is the calling convention you choose, and decisions you
make around compiling application expressions and definitions. Along with
your code, you will write a desgin document as a separate PDF describing how
your calling convention works. 

There are no restrictions on the calling convention you choose. You could
implement the x86-64 convention for your functions, the version we discussed
in class generalized to arbitrary lists of values, or something of your own
design. Different choices will require different implementation strategies,
different stack management requirements, and produce code that is
``interesting'' to debug in various ways.

In your design document, you should make sure to cover (in whatever order
makes sense) the items below. This report will be worth {\bf 30\% of your
assignment grade}.

\begin{enumerate}
  \item A description of your calling convention in general terms:
    \begin{enumerate}
      \item What does the caller do before and after the call?
      \item What is the callee responsible for?
      \item Are there improvements you can imagine making in the future?
    \end{enumerate}
    Use snippets of OCaml code from your compiler to illustrate key features.
  \item Pick three example programs that use functions and are interesting in
  different ways (you can use the required tests if you like), and use them
  to describe your calling convention:
  \begin{enumerate}
    \item Show their source, generated assembly, and output (you can
    summarize the generated code if it's quite long)
    \item Highlight the parts of the generated assembly that make the example
    interesting, distinct, and/or especially challenging to compile
  \end{enumerate}
\end{enumerate}

Still write this even if you don't have everything working! In that case, in
part 2, pick at least one example that {\it doesn't} work, and note both its
expected output and its actual behavior with your compiler.

Submit this part of the assignment as a PDF to {\tt pa4-written}.

{\bf Advice on writing}: Whenever you write, be thoughtful about your
expected audience, and most importantly {\it choose at least one specific
audience you are writing for}.\footnote{Writing for multiple audiences at the
same time is of course possible, hence ``at least'' one. It just requires
more care.} For this assignment, imagine that Diamondback is a custom
language that your team maintains. You are writing this design document to
help with the onboarding process for people joining your team. They are
programmers who know generally how compilers work and how x86-64 works, but
not how Diamondback has made its choices internally. They will be curious how
things work, why decisions were made the way they were, and what upcoming
work on the compiler might look like.


\section*{Rubric}

Note that half of the credit on this assignment is for testing and writing,
and the other half is for automatically-graded behavior:

\begin{itemize}
\item Required tests: 20\% (5 each)
\item Behavior of your compiler on our tests: 50\%
\item Design doc on calling convention: 30\%
\end{itemize}

\section*{Extensions}

These are not for credit, but you may enjoy thinking through, implementing,
or discussing in office hours or with your peers after the assignment is
complete.

\begin{enumerate}
\item Support tail calls.

\item In some cases, return type annotations can be omitted to reduce work on the
programmer. Make the return type annotations optional on function definitions,
and calculate them as part of building the definitions environment. Report an
error if the type cannot be calculated without the annotation.
\end{enumerate}

\end{document}